Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Basic separation model

We can make things a little more interesting by adding a heap to the mix, defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where \(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function. The Kleisli category of this monad can be equipped with the structure of a substructural effectful category by simply taking the affine/relevant morphisms to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\); as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we therefore have an \isotopessa model supporting the instructions \(\ms{get}: \mc{I}^\varnothing_0(\nats, \nats)\) and \(\ms{set}: \mc{I}^\varnothing_0(\nats \times \nats, ())\). We can assign semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} = \lambda (p, v)\;h. \{\iota_0((), [p \mapsto v]h, [])\}\). \(\ms{get}\) is a bit more tricky, since it is unclear what to do when we try to access uninitialized memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}} = \lambda p\;h. \{\iota_0(v, []) \mid v = h\;p \lor p \notin h\}\); for now, this will do.

@@

\TODO{by the above, this is an \isotopessa model w/ nice properties; moving on!}

Given a set \(A\), we can define a predicate \(\varphi\) on \(A\), \(\mb{Heap}\) pairs to have the following type:
\TODO{this}
\TODO{Objects \((A, \varphi)\), \((B, \psi)\)}
\TODO{arrows}
\TODO{note on magic wand}
\TODO{Tensor product and separating conjunction}
\TODO{Coproduct, Elgot structure inherited from base category}

\TODO{so, it's a substructural effectful category mmkay}
\TODO{central = no printing, no heap}
\TODO{linear objects iff non duplicable}
\TODO{relevant morphisms iff deterministic}
\TODO{affine morphisms always (if no heap effects?)}

\TODO{equality vs \cite{mellies-ftrs}}
\TODO{sliding w/ properly restricted equality}
\TODO{\(\implies\) central = no printing, heap allowed}

>>>>

We can make things a little more interesting by adding a heap to the mix, defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where \(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function. The Kleisli category of this monad can be equipped with the structure of a substructural effectful category by simply taking the affine/relevant morphisms to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\); as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we therefore have an \isotopessa model supporting the instructions \(\ms{set}: \mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}: \mc{I}^\varnothing_0(\nats, \nats)\), \(\ms{alloc}: \mc{I}^\varnothing_0(\nats, \nats)\), and \(\ms{free}: \mc{I}^\varnothing_0(\nats, \mb{1})\). We can assign semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} = \lambda (p, v)\;h. \{\iota_0((), [p \mapsto v]h, [])\}\). \(\ms{get}\) is a bit more tricky, since it is unclear what to do when we try to access uninitialized memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}} = \lambda p\;h. \{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\); for now, this will do. Finally, \(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', []) \mid h' = h \sqcup p \mapsto v\}\) simply fills a random empty heap cell with the provided value, returning a pointer to the cell, while \(\dnt{\ms{free}} = \lambda p\;h. \{\iota_0\;((), h \setminus p, [])\}\)

@@

Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of type \(A \to \ms{Heap} \to \ms{Prop}\). 
Taking as objects pairs \((A, \Phi)\), we can define the category \(\ms{Sep}\) to have morphisms
  \ms{Sep}((A, \Phi), (B, \Psi)) = \{f: A \to \ms{Comp}\;B 
    \mid \forall \phi: \ms{Heap} \to \ms{Prop}, \ms{hoare}\;f\;(\Phi \ast \phi)\;(\Psi \ast \phi)\}
where
  \ms{hoare}\;f\;\Phi\;\Psi = \forall a\;h. \Phi\;a\;h \implies \forall \iota_0\;(b, h') \in f\;a\;h. \Psi\;b\;h'
is the usual partial correctness triple definition and the separating conjunction of heap-predicates is defined as usual to be 
  \begin{gathered}
  \varphi \ast \psi = \lambda h. \exists h_1, h_2. h = h_1 \sqcup h_2 \land \varphi h_1 \land \psi h_2 
  \qquad
  \varphi \ast \Phi = \lambda a. \varphi \ast \Phi\;a
  \qquad
  \Phi \ast \varphi = \lambda a. \Phi\;a \ast \varphi \\
  \Phi \ast \Psi = \lambda (a, b). \Phi\;a \ast \Psi\;b: A \times B \to \ms{Heap} \to \ms{Prop}
  \end{gathered}
We can naturally define a tensor product on objects \((A, \Phi) \otimes (B, \Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking
  \item Monoidal unit \((\mb{1}, \ms{emp})\), where \(\ms{emp}\) is the predicate which holds only for the empty heap
  \item Associators and inverse associators given by \(\alpha_{(A, \Phi), (B, \Psi), (C, \Theta)} = \alpha_{A, B, C}\), which are valid since \(\ast\) is associative
  \item Unitors and inverse unitors given by \(\lambda_{(A, \Phi)} = \lambda_A\), \(\rho_{(A, \Phi)} = \rho_A\), which are valid since \(\ms{emp}\) is the unit of \(\ast\)
  \item \((C, \Theta) \otimes f = C \times f\) and \(f \otimes (C, \Theta) = f \times C\), which are valid definitions by the frame rule
equips \(\ms{Sep}\) with the structure of a premonoidal category. We can give \(\ms{Sep}\) the structure of a substructural effectful category by taking:
  \item Affine objects to be precisely those for which \(\ms{drop} = \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), which for this logic corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend on the heap.
  \item Relevant objects to be precisely those for which \(\ms{split} = \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), again in this case those for which \(\Phi\) does not depend on the heap.
  \item Central morphisms to be those which do not print \textit{or} modify the heap. We will consider how to relax this condition later.
  \item Relevant and affine morphisms to be the lifts of relevant/affine morphisms in \(\ms{Print}\)
\(\ms{Sep}\) also has coproducts given by \((A, \Phi) + (B, \Psi) = (A + B, [\Phi, \Psi])\), which allow us to equip it with an Elgot structure by taking advantage of the fact that
  f \in \ms{Sep}((A, \Phi), (B + A, [\Psi, \Phi])) \implies f^\dagger \in \ms{Sep}((A, \Phi), (B, \Psi))
This implies that \(\ms{Sep}\) is an \isotopessa model supporting the following interesting operations
  \item \(\ms{coe}: \mc{I}^\tint_1((A, \Phi), (A, \Psi))\) where \(\forall a, \Phi\;a \implies \Psi\;a\), implemented by the identity morphism
  \item \(\ms{print}: \mc{I}^\varnothing_0((\ms{Bytes}, \ms{emp}), (\mb{1}, \ms{emp}))\)
  \item \(\ms{rand}: \mc{I}^\taff_1((\mb{1}, \ms{emp}), (A, \ms{emp}))\) for \(A\) nonempty
  \item \(\ms{set}: \mc{I}^\varnothing_0((\nats \times \nats, \lambda (\ell, \cdot). \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\), where \(\ms{set}\) is modified slightly to return the address at which the value passed in was stored to overcome a lack of dependency between the source and target objects.
  \item \(\ms{get}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
  \item \(\ms{alloc}: \mc{I}^\varnothing_0((\nats, \ms{emp}), (\nats, \lambda \ell. \exists v. \ell \mapsto v))\)
  \item \(\ms{free}: \mc{I}^\varnothing_0((\nats, \lambda \ell. \exists v. \ell \mapsto v), (\mb{1}, \ms{emp}))\)

\TODO{equality vs \cite{mellies-ftrs}}
\TODO{sliding w/ properly restricted equality}
\TODO{\(\implies\) central = no printing, heap allowed: can make heap operations relevant/affine too!}
